Gravitational field calculation in spherical coordinates using variable densities in depth

### 2018-12-11 manuscript/manuscript.tex
Improve D and delta determination
Also update figures.

\citet{Uieda2016} compared the numerical integration of homogeneous density tesseroids
with the analytical solution of a spherical shell \citep{Mikuska2006,Grombein2013} in
order to obtain default values for the distance-size ratio $D$.
We will follow this idea but for our needs the spherical shell must
have the same density function of radius as our tesseroid model.
\citet{Lin2018} show the analytical solution of the gravitational potential generated by
a spherical shell with linear density in the radial coordinate.
We derive an expression for such gravitational potential in case the spherical shell has
an arbitrary density function in depth (Eq. \ref{eq:shell-pot}), and obtained the
expressions in case of linear and exponential density (see Appendix \ref{sec:shell}).

@@

\begin{table}
\caption{
    Description of the tesseroid models used to build a spherical shell and characterize
    the accuracy of the numerical integration.
    The first three columns describe the shapes of the tesseroid models: each one
    specifies the number of tesseroids on the corresponding dimension (radial,
    latitudinal, longitudinal).
    The last two columns specify the latitudinal and longitudinal size of each
    tesseroid that compose the spherical shell model.
    For each shape, we define several spherical shell models which the following
    thicknesses: 100m, 1km, 10km, 100km and 1000km.
    While the outer radius of every shell is equal to the mean Earth radius
}
\label{tab:tess-models}
\begin{tabular}{cccccc}
    Radial & Latitudinal & Longitudinal  & Total tesseroids & $L_\phi$ & $L_\lambda$
    \\ \hline
    1 & 1  & 2  & 2   & 180$^\circ$ & 180$^\circ$ \\
    1 & 6  & 12 & 72  &  30$^\circ$ &  30$^\circ$ \\
    1 & 12 & 24 & 288 &  15$^\circ$ &  15$^\circ$ \\
    1 & 18 & 36 & 648 &  10$^\circ$ &  10$^\circ$ \\
\end{tabular}
\end{table}

@@

The absolute density difference defined on equation
\ref{eq:density-abs-diff} will always be zero for the linear density case.
As a result, the inequality \ref{eq:delta-density} will always be satisfied and no
divisions will ever be performed.
Therefore, the distance-size ratio $D$ of the adaptive discretization algorithm is the
only mechanism that controls the accuracy of the numerical integration.
For this reason, we will only determine the minimum value of $D$ needed in order to
guarantee an acceptable accuracy while ignoring the value of $\delta$.

In order to compare the numerical results with the analytical solution we
must build a spherical shell made of tesseroids.
The outer radius of the shell is equal to the mean Earth radius and we repeat the
calculations for shell thicknesses of 1km and 35km.
We discretize the shell into a single layer of $30^\circ \times 30^\circ$ tesseroids and
compute their gravitational effects on $10 \times 10$ point grids with different
resolutions, heights, and locations on the sphere.
These grids are defined in Table \ref{tab:grids}.
The density of the shell and tesseroid model varies linearly with $r$
(Eq. \ref{eq:density-linear}) with angular coefficient

\begin{equation}
    a = -\frac{3300\text{kg/m$^3$} - 2670\text{kg/m$^3$}}{R - R_1},
\end{equation}

\noindent and linear coefficient

\begin{equation}
    b = \frac{3300\text{kg/m$^3$} -
        2670\text{kg/m$^3$}}{R - R_1} R +
        2670\text{kg/m$^3$},
\end{equation}

\noindent
in which $R$ = 6378.137 km is the mean Earth radius and $R_1$ is the inner radius of the
shell (determined by the thickness).

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{figures/linear-D.pdf}
\caption{
    Differences between the gravitational fields generated by the tesseroid model
    and the analytical solution for a thin (left) and a thick (right) spherical shell.
    Both models have a linear density function of radius (Eq. \ref{eq:density-linear}).
    The computations were performed on the four combinations described in
    Table \ref{tab:grids}.
    Due to the linearity of the density function, the density-based discretization
    algorithm is not applied.
    Differences are reported as a percentage of the shell values.
    The horizontal dashed black line represents a target difference of 0.1\%.
}
\label{fig:D-linear}
\end{figure*}

We compute the gravitational potential ($V$), the vertical component of the gradient
($g_z$), and the diagonal components of the Marussi tensor ($g_{xx}$, $g_{yy}$,
$g_{zz}$) for each grid in Table \ref{tab:grids}.
Other components are equal to zero outside of the shell and are thus omitted from the
analysis.
The computations are repeated for values of $D$ ranging from 0.5 to 10 with a step of
0.5.
We will then calculate the maximum absolute difference between these results and the
analytical solutions derived in Appendix \ref{sec:shell}.
The differences are shown in Fig. \ref{fig:D-linear} as relative to the spherical shell
values.
We omit the differences for $g_{xx}$ and $g_{yy}$ because they are equivalent to the
ones obtained for $g_{zz}$.
Finally, we set the optimal value of $D$ as the minimum value at which the corresponding
error of the numerical approximation is lower than 0.1\%.

>>>>

\citet{Uieda2016} compared the numerical integration of homogeneous density tesseroids
with the analytical solution of a spherical shell \citep{Mikuska2006, Grombein2013} in
order to obtain default values for the distance-size ratio $D$.
We will follow this idea but for our needs the spherical shell must
have the same density function of radius as our tesseroid model.
\citet{Lin2018} show the analytical solution of the gravitational potential generated by
a spherical shell with linear density in the radial coordinate.
We derive an expression for such gravitational potential in case the spherical shell has
an arbitrary density function in depth (Eq. \ref{eq:shell-pot}), and obtained the
expressions in case of linear and exponential density (see Appendix \ref{sec:shell}).

@@

\begin{table}
\caption{
    Description of the tesseroid models used to build a spherical shell and characterize
    the accuracy of the numerical integration.
    The first three columns describe the shapes of the tesseroid models: each one
    specifies the number of tesseroids on the corresponding dimension (radial,
    latitudinal, longitudinal).
    The last two columns specify the latitudinal and longitudinal size of each
    tesseroid that compose the spherical shell model.
    The outer radius of the shell is equal to the mean Earth radius, while the inner
    radius is defined based on its thickness.
    For each shape, we define several spherical shell models with the following
    thicknesses: 100m, 1km, 10km, 100km and 1000km.
}
\label{tab:tess-models}
\begin{tabular}{cccccc}
    Radial & Latitudinal & Longitudinal  & Total tesseroids & $L_\phi$ & $L_\lambda$
    \\ \hline
    1 & 1  & 2  & 2   & 180$^\circ$ & 180$^\circ$ \\
    1 & 6  & 12 & 72  &  30$^\circ$ &  30$^\circ$ \\
    1 & 12 & 24 & 288 &  15$^\circ$ &  15$^\circ$ \\
    1 & 18 & 36 & 648 &  10$^\circ$ &  10$^\circ$ \\
\end{tabular}
\end{table}

@@

The absolute density difference defined on equation
\ref{eq:density-abs-diff} will always be zero for the linear density case.
As a result, the inequality \ref{eq:delta-density} will always be satisfied and no
divisions will ever be performed.
Therefore, the two dimensional adaptive discretization algorithm is the
only mechanism that controls the accuracy of the numerical integration.
For this reason, we will only determine the minimum value of $D$ needed in order to
guarantee an acceptable accuracy while ignoring the value of $\delta$.

In order to compare the numerical results with the analytical solution we
must build spherical shell models made of tesseroids.
We propose the ones detailed on Table~\ref{tab:tess-models}: the outer radius of each
shell is equal to the mean Earth radius (6378.137 km) while the inner radius is
determined based on its thickness.
We cover spherical shells composed by tesseroid of different sizes and shells with
thicknesses of 100m, 1km, 10km, 100km and 1000km.

The comparison between the numerical results and the analytical solution could be
naively compared on a single computation point due to the rotation symmetry of the
spherical shell.
But the numerical results may adopt different values depending on the relative location
between the computation point and the tesseroid, thus is not wise to carry out the
comparison on a single point.
Instead, we propose four computation grids detailed on Table~\ref{tab:grids}: a Pole and
an Equator grids, and a Global and Satellite grids.
The former ones are regional scale grids located on the outer surface of the spherical
shell, while the latter are global scale grids located at zero height (Global) and at
the mean orbital height of the gravimetric satellite missions (Satellite).
We define these grids in the spirit of span a broad scenario of gravitational field
computations and ensure an acceptable accuracy on each one of them.
The comparison will be carried out by computing the maximum absolute difference between
the analytical solution and the numerical result on each point of the computation grid.

We assign a density that varies linearly with $r$ (Eq.~\ref{eq:density-linear}) to each
shell, which assumes the values of $\rho_\text{in} = 3300$~kg/m$^3$ and
$\rho_\text{out} = 2670$~kg/m$^3$ on the inner and outer radii, respectively.
So, the constants $a$ and $b$ from Eq.~\ref{eq:density-linear} must be

\begin{equation}
    a = \frac{\rho_\text{out} - \rho_\text{in}}{R_\text{out} - R_\text{in}},
\end{equation}

\begin{equation}
    b = \rho_\text{out} -
    \frac{\rho_\text{out} - \rho_\text{in}}{R_\text{out} - R_\text{in}} R_\text{out},
\end{equation}

\noindent
where $R_\text{out}$ and $R_text{in}$ are the outer and inner radii of the shell,
respectively.
For every shell defined in Table~\ref{tab:tess-models}, the outer radius will be equal
to the mean Earth radius (6378.137 km).

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{figures/linear-density-diffs.pdf}
\caption{
    Differences between the gravitational fields generated by each tesseroid shell model
    and the analytical solution.
    Each model has a linear density function of radius (Eq. \ref{eq:density-linear}).
    The computations were performed on the four combinations described in
    Table \ref{tab:grids}.
    Each line represents the maximum absolute difference between every shell model with
    a certain thickness.
    Due to the linearity of the density function, the density-based discretization
    algorithm is not applied.
    Differences are reported as a percentage of the shell values.
    The horizontal dashed black line represents a target difference of 0.1\%.
}
\label{fig:D-linear}
\end{figure*}

We compute the gravitational potential ($V$) and the vertical component of the gradient
($g_z$) that each shell model detailed on Table~\ref{tab:tess-models} generates on each
grid defined in Table~\ref{tab:grids}.
Other components of the gradient are equal to zero outside of the shell due to the
rotational symmetry and are thus omitted from the analysis.
The computations are repeated for values of $D$ ranging from 0.5 to 5 with a step of
0.5.
The differences are shown in Fig. \ref{fig:D-linear} as relative to the spherical shell
values.
We omit the differences for $g_{xx}$ and $g_{yy}$ because they are equivalent to the
ones obtained for $g_{zz}$.
Finally, we set the optimal value of $D$ as the minimum value at which the corresponding
error of the numerical approximation is lower than 0.1\%.
